
## 5. Fitting and choosing models

### 5.1 Back to the temperature series

```{r}
load("meteo.RData")
# plot data and trend:
plot(T.outside~date,meteo,type='l')
meteo$T.per = 18.2-4.9*sin(pi*(meteo$hours+1.6)/12)
lines(T.per~date,meteo,col='red')
```

### 5.2 Removing the diurnal periodicity
Assuming this is a sinus function, 
$$\alpha_1 + \alpha_2 \sin(t + \alpha_3)$$,
we need non-linear regression ($\alpha_3$)
```{r}
attach(meteo)
f = function(x) sum((T.outside - (x[1]+x[2]*sin(pi*(hours+x[3])/12)))^2)
nlm(f,c(0,0,0))
```

### 5.3 Temperature anomaly
```{r}
plot(T.outside-T.per~date, meteo, type='l')
title("anomaly")
```

### 5.4 What can we do with such models?
* try to find out which model fits best (model selection)
* learn how they were/could have been generated
* predict future observations (estimation/prediction/forecasting)
* generate similar data ourselves (simulation)

### 5.5 How to select a ``best'' model?
A possible approach is to find the minimum for {\em Akaike's Information
Criterion} (AIC) for ARMA($p,q$) models and series of length $n$:
$$AIC = \log \hat{\sigma}^2 + 2(p+q+1)/n$$
with $\hat{\sigma}^2$ the estimated residual (noise) variance.

Instead of finding a single best model using this single criterion, it
may be better is to select a small group of ``best'' models, and look
at model diagnostics for each: is the residual white noise? does it have
stationary variance?

Even better may be to keep a number of ``fit'' models and consider each as
(equally?) suitable candidates.

### 5.6 AIC for AR(p), and as as a function of $p$
```{r}
n = 10
aic = numeric(n)
for (i in 1:n)
   aic[i] = arima(T.outside, c(i,0,0))$aic  # AR(i)
aic
plot(aic[2:10], type='l')
```

### 5.7 Anomaly AIC as a function of $p$, for AR(p)
```{r}
an = T.outside - T.per
aic_an = numeric(n)
for (i in 1:n)
    aic_an[i] = arima(an,c(i,0,0))$aic  # AR(i)
aic_an
plot(aic_an[2:10], type='l')
```


```{r}
# Prediction, e.g. with AR(6)
x = arima(T.outside, c(6,0,0))

pltpred = function(xlim, Title) {
  plot(T.outside, xlim = xlim, type='l')
  start = nrow(meteo) + 1
  pr = predict(x, n.ahead = xlim[2] - start + 1)
  lines(start:xlim[2], pr$pred, col='red')
  lines(start:xlim[2], pr$pred+2*pr$se, col='green')
  lines(start:xlim[2], pr$pred-2*pr$se, col='green')
  title(Title)
}
pltpred(c(9860, 9900), "10 minutes")
pltpred(c(9400, 10000), "110 minutes")
pltpred(c(8000, 11330), "predicting 1 day")
pltpred(c(1, 19970), "predicting 1 week")

plot(T.outside,xlim=c(1,19970), type='l')
x.an = arima(an, c(6,0,0)) # model the anomaly by AR(6)
x.pr = as.numeric(predict(x.an, 10080)$pred) 
x.se = as.numeric(predict(x.an, 10080)$se)
hours.all = c(meteo$hours, max(meteo$hours) + (1:10080)/60)
T.per = 18.2-4.9*sin(pi*(hours.all+1.6)/12)
lines(T.per, col = 'blue')
hours.pr = c(max(meteo$hours) + (1:10080)/60)
T.pr = 18.2-4.9*sin(pi*(hours.pr+1.6)/12)
lines(9891:19970, T.pr+x.pr, col='red')
lines(9891:19970, T.pr+x.pr+2*x.se, col='green')
lines(9891:19970, T.pr+x.pr-2*x.se, col='green')
title("predicting 1 week: periodic trend + ar(6) for anomaly")

x = arima(T.outside, c(6,0,0))
plot(T.pr + arima.sim(list(ar = x.an$coef[1:6]), 10080, sd = sqrt(0.002556)), 
	col = 'red', ylab="Temperature")
lines(18+arima.sim(list(ar = x$coef[1:6]), 10080, sd=sqrt(0.002589)), 
	col = 'blue')
title("red: with trend, blue: without trend")
```

### 5.8 What can we learn from this?}
Prediction/forecasting:
* AR(6) prediction is a compromise between the end of the series and the trend
* the closer we are to observations, the more similar the prediction is to the
nearest (last) observation
* further in the future the prediction converges to the trend
* the more useful (realistic) the trend is, the more realistic 
the far-into-the-future prediction becomes
* the standard error of prediction increases when predictions are further in the
future.

### 5.9 Next steps: how do we fit coefficients?
