# Time Series

## Intro: How do R markdown / quarto files work?

* load them with rstudio
* click the `knit` or `render` button
* debug/step through:
    * load all code up to here-button
    * ctrl-Enter: run this line
    * output in console, objects appear in Environment browser


## Time series models

we will first look into time series models, because they are
+ simple
+ easy to write down
+ well understood

time series models are roughly divided in 

1. time domain models and, which look at correlations and memory
2. frequency domain models, which focus on periodicities 

Spatial equivalents are mostly found in (a), although (b) has
spatial equivalences as well (e.g. wavelets).

### some data

Consider the following process ($\Delta t$  = 1 min):

```{r fig.width=10, fig.height=5}
load("meteo.RData") # should be available in the current working directory
ls()
names(meteo)
plot(T.outside~date, meteo, type='l', ylab = parse(text = "Temperature ({}*degree* C)"), xlab = "date, 2007")
title("Outside temperature, Hauteville, FR")
```

### Questions
+ how can we describe this process in statistical terms?
+ how can we model this process?
+ (how) can we predict future observations?

### White noise, and AR($n$)
Perhaps the simplest time series model is _white noise_ with mean $m$:

$$y_t = m + e_t, \ \ e_t \sim N(0,\sigma^2)$$

$N(0,\sigma^2)$ denoting the normal distribution with mean 0 and
variance $\sigma^2$, and $\sim$ meaning _distributed as_ or _coming from_.

$t$ is the index $t=1,2,...,n$ of the observation, and refers to
specific times, which, when not otherwise specified are at regular
intervals.

A *white noise* process is completely without memory: each observation is
independent from its past or future. Plotting independent, standard normal
values against their index (the default for plotting a vector in R) shows
how a white noise time series would look like:

```{r fig.width=10, fig.height=5}
white.noise = rnorm(100)
plot(white.noise, type='b')
title("100 independent observations from N(0,1)")
```

```{r fig.width=10, fig.height=5}
white.noise = rnorm(1000)
plot(white.noise, type='l')
title("1000 independent observations from N(0,1)")
```

```{r fig.width=10, fig.height=5}
white.noise = rnorm(10000)
plot(white.noise, type='l')
title("10000 independent observations from N(0,1)")
```

We can look at the auto-correlation function of a white noise
process, and find it is uncorrelated for any lag larger than 0:

```{r fig.width=10, fig.height=5}
plot(acf(white.noise))
```

Correlation x and y:
$$
r(x,y) = \frac{\sum_{i=1}^{n}(x_i-\bar{x})(y_{i}-\bar{y})}{\sqrt{\sum_{i=1}^n (x_i-\bar{x})^2\sum_{i=1}^n(y_i-\bar{y})^2}}
$$

### Autocorrelation
Autocorrelation (or lagged correlation) is the correlation between $y_i$ and $y_{i+h}$, as a function of the lag $h$:
$$
r(h) = \frac{\sum_{i=1}^{n-h}(y_i-\bar{y})(y_{i+h}-\bar{y})}{\sum_{i=1}^n (y_i-\bar{y})^2}
$$
with $\bar{y} = \frac{1}{n} \sum_{i=1}^n y_i$

### Random walk
A simple, next model to look at is that of _random walk_, where each
time step a change is made according to a white noise process:
$$y_t = y_{t-1} + e_t$$
Such a process has memory, and long-range correlation. If we take the first-order
differences,
$$y_t - y_{t-1} = e_t$$
we obtain the white noise process.

Further, the variance of the process increases with increasing domain
(i.e., it is non-stationary)

### Example random walk:
We can compute it as the cumulative sum of standard normal deviates: $y_n = \sum_{i=1}^n
e_i$:

```{r fig.width=10, fig.height=5}
# generate three series:
rw1 = cumsum(rnorm(5000))
rw2 = cumsum(rnorm(5000))
rw3 = cumsum(rnorm(5000))
plot(rw1, type='l', ylim = range(c(rw1,rw2,rw3)))
lines(rw2, type='l', col='red')
lines(rw3, type='l', col='blue')
```

```{r fig.width=10, fig.height=5}
plot(acf(rw3))
```

```{r fig.width=10, fig.height=5}
plot(acf(diff(rw3)))
```

### MA(1), MA(q)
Let $e_t$ be a white noise process. A moving average process of
order $q$ is generated by
$$y_t = \beta_0 e_t + \beta_1 e_{t-1} + ... + \beta_q e_{t-q}$$

Note that the $\beta_j$ are weights, and could be $\frac{1}{q+1}$ to
obtain an unweighted average. Moving averaging smoothes the white
noise series $e_t$.

Moving average over monthly CO2 measurements on Maunaloa:
```{r fig.width=10, fig.height=5}
plot(co2)
lines(filter(co2, rep(1/12, 12)), col='blue')
```

Moving averages over a white noise process:
```{r fig.width=10, fig.height=5}
#set.seed(13531)
e = rnorm(2000)
plot(e, type='l')
e5 = filter(e, rep(1/5, 5))
e20 = filter(e, rep(1/20, 20))
lines(e5, col='red')
lines(e20, col='blue')
```
```{r}
acf(e)
acf(e5, na.action = na.pass)
acf(e20, na.action = na.pass)
```

Wider moving average filters give new processes with
+ less variation
+ stronger correlation, over larger lags

### AR(1), AR(p)
An auto-regressive (1) model, or AR(1) model is generated by
$$y_t = \phi_1 y_{t-1}+e_t$$
and is sometimes called a Markov process. Given knowledge of $y_{t-1}$,
observations further back carry no information; more formally:
$$\Pr(y_t|y_{t-1},y_{t-2},...,y_{t-q}) = \Pr(y_t|y_{t-1})$$

+ $\phi_1 = 1$ gives random walk, $\phi_1=0$ gives white noise.
+ AR(1) processes have correlations beyond lag 1
+ AR(1) processes have non-significant _partial autocorrelations_ beyond lag 1

As an example, we create (simulate) an AR(1) process with $\phi_1=0.85$ and $e$
drawn from the standard normal distribution (mean 0, variance 1).

```{r}
n = 1000
y = rep(NA_real_, n) # initialise the variable: not needed, but good practice
y[1] = 0
for (i in 2:n) y[i] = 0.85 * y[i-1] + rnorm(1)
plot(y, type = 'l', main = "AR(1)")
plot(acf(y))
```

Compare the shape of the `acf` with that obtained from a MA process: different!

### AR(p)
$$y_t = \phi_1 y_{t-1}+ \phi_2 y_{t-2} + ... + \phi_p y_{t-p} + e_t$$
or
$$y_t = \sum_{j=1}^p \phi_j y_{t-j}+e_t$$
+ The state of $y_t$ does not _only_ depend on $y_{t-1}$, but
observations further back contain information
+ AR(p) have autocorrelations beyond lag p
+ AR(p) have "zero" _partial_ autocorrelations beyond lag p

```{r}
y2 = rep(NA_real_, 1000) # initialise the variable: not needed, but good practice
y2[1] = 0; y2[2] = rnorm(1)
for (i in 3:1000) y2[i] = 0.5 * y2[i-1] + 0.15 * y2[i-2] + rnorm(1)
plot(y2, type = 'l', main = "AR(2)")
plot(acf(y2))
```


### Partial correlation
+ Correlation between $y_t$ and $y_{t-2}$ is simply
obtained by plotting both series of length $n-2$, and computing 
correlation
+ Lag-2 _partial_ autocorrelation of $y_t$ and $y_{t-2}$, given
the value inbetween $y_{t-1}$ is obtained by
    + computing residuals $\hat{e}_t$ from regressing of $y_t$ on $y_{t-1}$
    + computing residuals $\hat{e}_{t-2}$ from regressing of $y_{t-2}$ on $y_{t-1}$
    + computing the correlation between both residual series $\hat{e}_t$ and
 $\hat{e}_{t-2}$.
+ Lag-3 partial autocorrelation regresses $y_t$ and $y_{t-3}$ 
on _both_ intermediate values $y_{t-1}$ and $y_{t-2}$
+ etc.

Partial correlation can help reveal what the order of an MA(q) or AR(p) series is:

```{r}
plot(pacf(e5, na.action = na.omit))
plot(pacf(e20, na.action = na.omit))
plot(pacf(y))
plot(pacf(y2))
```

### Relation between AR and MA processes
Chatfield [@chatfield] has more details about this.  Substitute the AR(1) as follows
$$y_t = \phi_1 y_{t-1} + e_t$$
$$y_t = \phi_1 (\phi_1 y_{t-2} + e_{t-1}) + e_t$$
$$y_t = \phi_1^2 (\phi_1 y_{t-3} + e_{t-2}) + \phi_1 e_{t-1} + e_t$$
etc. In the limit, we can write any AR process as an (infinite)
MA process, and vice versa.
